\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage[numbers]{natbib}
\usepackage{tikz}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\geometry{
  left=2.5cm,
  right=2.5cm,
  top=2.5cm,
  bottom=2.5cm
}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

\title{\textbf{NRR-Projection: Projection Retention, State Reduction, and Identifiability in Non-Resolution Reasoning}}

\author{
  Kei Saito\thanks{ORCID: 0009-0006-4715-9176} \\
  Independent Researcher, Japan \\
  \texttt{kei.saito.research@gmail.com}
}

\date{March 2026 \\[0.5em]
\textit{Part of the Non-Resolution Reasoning (NRR) research program.}}

\begin{document}
\maketitle

\begin{center}
\textcopyright\ 2026 Kei Saito.
Licensed under CC BY 4.0.\\
\url{https://creativecommons.org/licenses/by/4.0/}
\end{center}

\begin{abstract}
NRR-Phi \cite{saito2026phi} defined a text-to-state mapping in which multiple admissible interpretations can remain jointly represented rather than being eliminated at the first output step. This paper asks a narrower theoretical question: under what conditions can retaining multiple admissible projections preserve more state-distinguishing structure than reducing analysis to a single retained view? We answer this question by taking the NRR state defined in Phi as the default object of analysis and adding projection-family machinery for the comparison.

We distinguish non-destructive output projection, which emits one interpretation while preserving internal state, from destructive collapse, which removes non-selected alternatives from the retained state. Under a task-conditioned distribution on admissible states, we show that a retained family of projections is at least as informative about the state as any one of its member projections alone. We additionally show that when a projection family is separating on the admissible state set, it identifies distinctions that any non-injective single projection cannot preserve by itself. Read in the analytical correspondence where destructive reduction is treated as retention of a single view, these results provide a condition-bounded explanation for why premature reduction can be information-disadvantageous.

To connect the theory to existing evidence without overclaiming, we reinterpret already reported NRR-Boundary findings \cite{saito2026boundary}. In particular, provider-by-order sign flips and conditional-first metric-dependent differences can be read as differences in retained view order and state-flattening behavior. This section is explicitly post hoc and is not presented as a confirmatory test of the theory.

NRR is not an anti-LLM framework. NRR does not replace standard LLM use. In the narrower fixed-evidence setting studied here, the practical issue is whether output can remain state-preserving long enough to avoid premature destructive reduction.
\end{abstract}

\noindent\textbf{Keywords:} non-resolution reasoning, projection, identifiability, state reduction, information retention, conditional utility

\section{Introduction}
\label{sec:intro}

NRR-Core \cite{saito2025nrr} and NRR-Phi \cite{saito2026phi} argue that premature fixation can remove useful structure from reasoning trajectories. NRR-Phi \cite{saito2026phi} formalized this idea by defining a text-to-state mapping in which multiple admissible interpretations remain represented in a single state rather than being collapsed immediately at output time. What remains underdeveloped is the narrower theoretical question of \emph{why} retaining non-resolved structure can be information-advantageous under some conditions.

This paper addresses that question without introducing a new metaphysics of ``hidden meaning.'' The default object of analysis is the NRR state itself. The target claim is correspondingly narrow: reducing analysis to one retained view can remove distinctions that a retained family of admissible projections would preserve, and destructive collapse is interpreted through that single-view reduction lens. This is not a claim that unresolved states are always better, nor that ambiguity is inherently desirable. It is a condition-bounded claim about information retention, identifiability, and the costs of premature state reduction.

The motivating intuition is familiar from multi-view observation. A single view may flatten distinctions that become recoverable once multiple admissible views are retained together \cite{hartley2004multiple}. In this manuscript that intuition is only motivational. The formal results concern state representations and projection families, not a full claimed identity between semantic interpretation and geometric vision.

NRR is not an anti-LLM framework.
NRR does not replace standard LLM use.
NRR optimizes when to commit and when to defer, under explicit conditions.

NRR is a derivation-and-evaluation framework, not a closed recipe.
The projection family analyzed here is an evaluated family under the stated conditions, not an exhaustive set; extension claims require the same fixed protocol and condition-bounded reporting.
Within the NRR series, this paper is positioned as a companion line to NRR-Phi and a bridge line linking the Phi state ontology to the empirical boundary patterns reported in NRR-Boundary.

\textbf{Series Consistency Statement:}
Unsupported interpretations are never injected at a fixed evidence state, hypothesis-set expansion is evidence-gated, and unobserved possibilities remain explicitly unresolved. In this paper, the evidence state is fixed before analysis, admissible views are taken only over evidence-consistent states, and no gain claim relies on adding unsupported alternatives.

\subsection{Contributions}

\begin{enumerate}
  \item A Phi-grounded formalization that distinguishes non-destructive output projection from destructive collapse.
  \item A theorem showing information monotonicity for retained projection families under a task-conditioned distribution on admissible states.
  \item A theorem showing identifiability gain when the retained projection family is separating on the admissible state set.
  \item A state-reduction remark making explicit that destructive collapse can erase alternative-sensitive structure needed by later operations.
  \item A cost-bounded remark clarifying why the framework implies conditional utility rather than universal superiority.
  \item A post hoc reinterpretation of existing NRR-Boundary results \cite{saito2026boundary} as an empirical contact point for the theory.
\end{enumerate}

\section{Relation to NRR-Phi}
\label{sec:phi}

NRR-Phi \cite{saito2026phi} defines a semantic state instance as
\begin{equation}
S = \{(v_i, c_i, w_i, m_i)\}_{i=1}^{n},
\end{equation}
where multiple admissible interpretations may remain jointly represented. The present paper adopts that state as its default object of analysis rather than introducing a separate privileged semantic object.

This choice matters because projection and collapse are not equivalent in Phi. Phi already permits singular output without destructive internal reduction. We therefore distinguish two operations.

\begin{definition}[Non-destructive output projection]
For an NRR state $S$, a non-destructive output projection $\Pi_j$ emits one output view $y_j$ while retaining the underlying state:
\begin{equation}
\Pi_j(S) = (y_j, S).
\end{equation}
\end{definition}

\begin{definition}[Destructive collapse]
For an NRR state $S$, a destructive collapse operator $C_j$ retains one interpretation-bearing component and removes the non-selected alternatives from the retained state:
\begin{equation}
C_j(S) = S_j.
\end{equation}
\end{definition}
\begin{figure}[t]
\centering
\begin{tikzpicture}[
  x=1cm,
  y=1cm,
  >=stealth,
  every node/.style={font=\small},
  state/.style={draw, rectangle, minimum width=3.0cm, minimum height=1.6cm, align=center},
  outbox/.style={draw, rectangle, minimum width=1.8cm, minimum height=1.0cm, align=center}
]
\node[state] (S) at (0,0) {$S$\\$s_1, s_2, \ldots, s_n$};
\coordinate (u) at (2.1,0.6);
\coordinate (d) at (2.1,-0.6);
\node[outbox] (y) at (4.1,1.4) {$y_j$};
\node[state] (Skeep) at (7.4,1.4) {$S$\\$s_1, s_2, \ldots, s_n$};
\node[state] (Sj) at (6.0,-1.7) {$S_j$\\$s_j$ only};

\draw[->] (S.east) -- (u) node[midway, above] {$\Pi_j$};
\draw[->] (u) -- (y.west);
\draw[->] (u) -- (Skeep.west);
\node[align=center] at (5.7,0.6) {\footnotesize internal state retained};

\draw[->] (S.east) -- (d) node[midway, below] {$C_j$};
\draw[->] (d) -- (Sj.west);
\node[align=center] at (4.7,-0.9) {\footnotesize alternatives removed};
\end{tikzpicture}
\caption{Distinction between non-destructive output projection $\Pi_j$ and destructive collapse $C_j$. Both operate on the same NRR state $S$, but only $C_j$ removes non-selected alternatives from the retained state.}
\label{fig:projection_vs_collapse}
\end{figure}

The theoretical comparisons in this paper are about retained multi-view structure versus the single-view analytical regime associated with destructive collapse $C_j$, not about NRR versus $\Pi_j$. The latter is already compatible with Phi because internal alternatives remain available after output.

\section{Formal Setup}
\label{sec:setup}

\subsection{Evidence state and admissible state set}

Fix an evidence state $E$. Let $\mathcal{S}_E$ denote the finite set of admissible NRR states consistent with that evidence state. All views, reductions, and comparisons below are restricted to $\mathcal{S}_E$. This restriction prevents gain claims from depending on unsupported interpretation injection.

For a retained component index $j$, define the reduced-state space
\begin{equation}
\mathcal{S}_E^{(j)} := \{S_j : S \in \mathcal{S}_E\},
\end{equation}
that is, the set of states obtained by collapsing admissible full states to their $j$th retained component.

\subsection{Projection family}

\begin{definition}[Admissible projection family]
Let $\{P_a\}_{a \in A}$ be a family of admissible projection operators on $\mathcal{S}_E$, where each
\begin{equation}
P_a : \mathcal{S}_E \to \mathcal{Y}_a
\end{equation}
returns a partial view of the state.
\end{definition}

Examples include one interpretation choice, one contextual framing, one order-conditioned readout, or one output rendering of the current state. The framework does not require every such view to be equally useful; it requires only that these views be defined over the same admissible state set.

\subsection{Probabilistic reading}

For the information-theoretic result below, we treat $S$ as a random variable distributed over $\mathcal{S}_E$ under a task-conditioned probability measure $\mu_E$. This measure is not part of the ontology of NRR-Phi \cite{saito2026phi} itself; it is an analytical device for comparing retained views. Since $\mathcal{S}_E$ is finite, all mutual information quantities below are well-defined and finite. If this probabilistic setup is considered heavier than needed in a final version, the identifiability theorem in Section~\ref{sec:ident} can stand alone as the primary result.

\section{Main Results}
\label{sec:results}

The connection between the formal setup and the collapse operators defined in Section~\ref{sec:phi} is as follows. Destructive collapse $C_j$ retains only one interpretation-bearing component of $S$. In the analytical framework below, this corresponds to retaining a single projection view $Y_1 = P_a(S)$. Non-destructive retention of the full state preserves access to the family $Y_k = (P_{a_1}(S), \ldots, P_{a_k}(S))$. The operators $\Pi_j$ defined in Section~\ref{sec:phi} and the projection family $\{P_a\}_{a \in A}$ serve different roles: $\Pi_j$ is a system-level operation that emits an output while preserving internal state, whereas $P_a$ is an analytical readout used to formalize the comparison between retained and reduced views. The theorems below concern the information content of these analytical readouts, not the emission mechanism.

\begin{definition}[Collapse-induced projection]
\label{def:collapse_proj}
For each destructive collapse operator $C_j$, define the collapse-induced projection $P_j^{\mathrm{coll}} : \mathcal{S}_E \to \mathcal{S}_E^{(j)}$ by
\begin{equation}
P_j^{\mathrm{coll}}(S) := C_j(S) = S_j.
\end{equation}
\end{definition}

Theorems~\ref{thm:monotonicity} and~\ref{thm:ident} hold for any admissible projection family. The application to destructive collapse is the special case obtained by taking the projection family to include collapse-induced projections $\{P_j^{\mathrm{coll}}\}$. In that case, retaining one collapse-induced projection corresponds, as a static observational comparison, to retaining the output of one destructive collapse operator. The general results therefore apply to this comparison; they characterize information content of the retained readout, not the full dynamic consequences of state overwriting.

\subsection{Information monotonicity}

\begin{theorem}[Retained-family monotonicity]
\label{thm:monotonicity}
Let
\begin{equation}
Y_k = (P_{a_1}(S), \ldots, P_{a_k}(S))
\end{equation}
and let
\begin{equation}
Y_1 = P_{a_j}(S)
\end{equation}
for one selected retained view from the same family. Then
\begin{equation}
I(S;Y_k) \geq I(S;Y_1).
\end{equation}
\end{theorem}

\begin{proof}
$Y_1$ is a deterministic function of $Y_k$ obtained by selecting one coordinate. By the data processing inequality \cite{cover2006elements}, mutual information cannot increase under deterministic post-processing, so
\begin{equation}
I(S;Y_k) \geq I(S;Y_1).
\end{equation}
\end{proof}

The theorem does not claim that each additional retained view improves downstream utility in every operational setting. It states only that a retained family cannot be less informative about the state than one of its own retained members alone, before cost terms are introduced.
The contribution of Theorem~\ref{thm:monotonicity} is therefore not the novelty of the inequality itself, but the fact that this comparison is formally safe once projection and destructive collapse are separated inside the Phi ontology.

\subsection{Identifiability under separating families}
\label{sec:ident}

\begin{definition}[Separating family]
A projection family $\{P_a\}_{a \in A}$ is separating on $\mathcal{S}_E$ if for any distinct states $S \neq S'$ there exists some $a \in A$ such that
\begin{equation}
P_a(S) \neq P_a(S').
\end{equation}
\end{definition}

\begin{theorem}[Identifiability gain]
\label{thm:ident}
If $\{P_a\}_{a \in A}$ is separating on $\mathcal{S}_E$, then the full retained family identifies distinctions on $\mathcal{S}_E$ that any non-injective single projection fails to identify by itself.
\end{theorem}

\begin{proof}
If the family is separating, then for any distinct states $S \neq S'$ at least one family member distinguishes them. Hence the map
\begin{equation}
S \mapsto (P_a(S))_{a \in A}
\end{equation}
is injective on $\mathcal{S}_E$. Any single non-injective projection $P_b$ admits distinct states $S \neq S'$ with $P_b(S)=P_b(S')$, so that projection alone cannot identify those distinctions.
\end{proof}

This is the structural core of the argument. When destructive collapse is read analytically as retention of one view, it can flatten distinctions that a sufficiently rich retained family preserves. In that limited sense, the theorem matches a broader multi-view intuition: multiple coordinated views can preserve distinctions that single-view access misses under appropriate structural conditions.

\subsection{Reduction loss inside Phi}

Even without the probabilistic reading of Theorem~\ref{thm:monotonicity}, a weaker but still useful statement remains.

\begin{remark}[State-reduction loss]
\label{rem:reduction}
Let $C_j : \mathcal{S}_E \to \mathcal{S}_E^{(j)}$ be a destructive collapse operator, and assume $\mathcal{S}_E$ contains at least one state with two or more relevant non-identical alternatives. Let $F$ be a functional defined on a common representation that accommodates both full states in $\mathcal{S}_E$ and reduced states in $\mathcal{S}_E^{(j)}$ (for example, the set of all finite interpretation sets under set-level operations). If $F$ depends on alternative-sensitive structure, then there exist states such that
\begin{equation}
F(S) \neq F(C_j(S)).
\end{equation}
This is definitionally immediate from the structure of $C_j$, but is stated explicitly to make clear that destructive collapse can erase structure that may be needed by later operations.
\end{remark}

\subsection{Conditional utility}

\begin{remark}[Cost-bounded utility]
\label{rem:utility}
Let downstream utility be
\begin{equation}
U = G(\text{information retained}) - \lambda \cdot \text{Cost}(\text{retention}),
\end{equation}
with $\lambda > 0$. If $G$ is increasing in retained information and $\text{Cost}$ is increasing in the number of retained views, then the sign of the utility difference depends on the operating regime. This paper does not prove the existence of both retention-favoring and collapse-favoring regimes as a theorem; it observes that the framework is structurally compatible with both, which is necessary for consistency with the NRR guardrail that non-resolution is conditionally useful rather than universally superior. The form of this tradeoff is also consistent with information bottleneck style reasoning, where retained information is balanced against compression cost rather than treated as unconditionally desirable \cite{tishby1999information}.
\end{remark}

\section{Boundary Reinterpretation}
\label{sec:boundary}

The preceding results are mathematical and do not require experiment to stand. However, theory-only presentation can look disconnected from the existing NRR empirical line. To provide a limited empirical contact point, this section reinterprets already reported NRR-Boundary findings \cite{saito2026boundary} through the projection framework.

\textbf{Important scope note:} this section is a post hoc reinterpretation of existing results, not a confirmatory test of NRR-Projection. It shows only that the framework can organize already observed empirical regularities. No formal mapping between the Boundary operators (\texttt{conditional}, \texttt{branch}) and specific projection family members $P_a$ is defined in this paper. The readings below are therefore analogical: they show that the Projection vocabulary can organize the observed patterns, not that the patterns deductively follow from the theorems. A direct empirical test would require a separately fixed Projection-specific design.

\subsection{Order-sensitive sign flips}

NRR-Boundary \cite{saito2026boundary} reported ordered-combination cases in which the same operator set changed sign when the order changed. In particular, the pattern \texttt{branch\_conditional} showed sign-flip boundaries on several metrics, with Gemini moving opposite to Anthropic/OpenAI on rows including \texttt{irp\_monotonic\_progress\_rate}, \texttt{misinterpretation\_rate}, \texttt{rework\_cost\_tokens}, and \texttt{stability\_variance}.

These values are reproduced from existing Boundary results, not from new experiments conducted for this paper.

\begin{table}[h]
\centering
\small
\begin{tabular}{p{5.2cm}ccc}
\toprule
Metric & Anthropic sign & OpenAI sign & Gemini sign \\
\midrule
irp\_monotonic\_progress\_rate & $-$ & $-$ & $+$ \\
misinterpretation\_rate & $+$ & $+$ & $-$ \\
rework\_cost\_tokens & $+$ & $+$ & $-$ \\
stability\_variance & $+$ & $+$ & $-$ \\
\bottomrule
\end{tabular}
\caption{Sign-flip cases for \texttt{branch\_conditional}. Data from NRR-Boundary \cite{saito2026boundary}; reproduced from existing Boundary results for post hoc reinterpretation, not from new experiments conducted for this paper. Signs denote metric-aligned improvement direction in the Boundary summaries ($+$ = improvement, $-$ = degradation), not raw metric increase/decrease. Signs reflect rep3 aggregation; rep1 shows zero for some OpenAI rows on these metrics. Source file: \texttt{stats/combo\_rep3\_all/mst\_sign\_flip\_boundaries.csv} in the NRR-Boundary repository.}
\label{tab:boundary_signflip}
\end{table}

A compatible projection-side reading is as follows. Order changes which partial view is retained earlier in the trajectory. If a provider's internal behavior makes one early retained view flatten the admissible state more aggressively than another, then downstream quality metrics can reverse direction across providers without any contradiction in the theory. The theory does not predict a universal preferred order; it predicts that different retained-view orders can preserve or destroy different distinctions.

\subsection{Conditional-first metric-specific improvements}

NRR-Boundary \cite{saito2026boundary} also reported that \texttt{conditional\_branch} showed no sign-flip boundary on several practical metrics. At the same time, it showed a negative provider-balanced improvement mean on \texttt{irp\_monotonic\_progress\_rate} ($-0.0513$), so the profile discussed here is metric-specific rather than uniformly favorable. It also showed provider-balanced improvements on:
\begin{itemize}
  \item \texttt{irp\_oscillation\_rate}
  \item \texttt{misinterpretation\_rate}
  \item \texttt{rework\_cost\_tokens}
  \item \texttt{stability\_variance}
\end{itemize}

These values are reproduced from existing Boundary results, not from new experiments conducted for this paper.

\begin{table}[h]
\centering
\small
\begin{tabular}{p{4.3cm}cc}
\toprule
Metric & Provider-balanced improvement (mean) & Provider-balanced direction \\
\midrule
irp\_oscillation\_rate & 0.0880 & positive \\
misinterpretation\_rate & 0.0571 & positive \\
rework\_cost\_tokens & 85.5 & positive \\
stability\_variance & 0.0153 & positive \\
\bottomrule
\end{tabular}
\caption{Conditional-first metric-specific cases for \texttt{conditional\_branch} with provider-balanced improvement means. Data from NRR-Boundary \cite{saito2026boundary}; reproduced from existing Boundary results for post hoc reinterpretation, not from new experiments conducted for this paper. Source file: \texttt{stats/combo\_rep3\_all/mst\_tau\_trace\_provider\_balanced.csv} in the NRR-Boundary repository.}
\label{tab:boundary_conditional_first}
\end{table}

A compatible projection-side reading, again condition-bounded, is as follows. If the conditional view retains discriminative structure that later branching decisions need, then preserving that view earlier can reduce downstream flattening and thereby improve later identification and update stability on some metrics, while still leaving other metrics unchanged or worse. This is not a proof of the projection theory. It is a compatible reading of already observed evidence.
This does not elevate conditional-first ordering into a general rule; the metric-specific improvement observed here is specific to the metrics and providers in the Boundary dataset.

\section{Limits and Non-Claims}
\label{sec:limits}

This paper makes several explicit non-claims.

\begin{itemize}
  \item It does not claim that unresolved states are always better.
  \item It does not claim that ambiguity itself is inherently good.
  \item It does not claim that visual parallax and semantic non-resolution are fully identical systems.
  \item It does not claim that entropy alone measures practical usefulness.
  \item It does not claim that unsupported interpretations may be added for the sake of richness.
  \item It does not equate non-destructive output projection with destructive collapse.
\end{itemize}

Theorems~\ref{thm:monotonicity} and~\ref{thm:ident} show structural facts about retained view families and state distinguishability. They do not, by themselves, determine the best operational policy for every system. That choice depends on cost, latency, memory constraints, and the risk associated with delayed versus premature commitment.
The separating condition in Theorem~\ref{thm:ident} is a sufficient condition for identifiability, not a claim that separating families are routinely available in all operational settings.

The Boundary reinterpretation section is especially limited. Because it is post hoc, it cannot serve as a confirmatory test of the theory. A future empirical test would need to manipulate retained-view number, retained-view order, or collapse timing under a fixed protocol designed in advance for Projection-specific hypotheses.

\section{Conclusion}
\label{sec:conclusion}

This paper isolates one narrow theoretical contribution within the NRR program. Starting from the state ontology defined by NRR-Phi \cite{saito2026phi} and adding an analytical framework of projection families and task-conditioned distributions, it shows why a retained family of admissible projections can preserve distinctions that a single retained view does not. Read through the analytical correspondence developed in this paper, this clarifies why premature destructive reduction can be costly under some conditions. The core result is not that non-resolution is universally superior. The result is that information retention and identifiability can favor non-destructive retention under some conditions, while other regimes can rationally favor collapse.

The practical role of this paper is therefore foundational rather than replacement-oriented. NRR-Core \cite{saito2025nrr} defined the program, NRR-Phi \cite{saito2026phi} defined the state machinery, and NRR-Projection explains one reason why premature destructive reduction can be costly. Existing Boundary results \cite{saito2026boundary} provide a compatible empirical contact point, but a direct confirmatory test of the present theory remains future work. This is not an ordering-recommendation paper; the Boundary connection is interpretive rather than prescriptive.

\begin{thebibliography}{9}

\bibitem{saito2025nrr}
K.~Saito.
\newblock NRR-Core: Non-Resolution Reasoning as a Computational Framework for Contextual Identity and Ambiguity Preservation.
\newblock arXiv preprint arXiv:2512.13478, 2025.

\bibitem{saito2026phi}
K.~Saito.
\newblock NRR-Phi: Text-to-state mapping for ambiguity preservation in LLM inference.
\newblock arXiv preprint arXiv:2601.19933, 2026.

\bibitem{saito2026boundary}
K.~Saito.
\newblock NRR-Boundary: Same operators, opposite effects by provider and order.
\newblock Available at \url{https://github.com/kei-saito-research/nrr-boundary}. Manuscript and data publicly available, 2026. Commit \texttt{4cbb80c} accessed March 2026.

\bibitem{cover2006elements}
T.~M. Cover and J.~A. Thomas.
\newblock \emph{Elements of Information Theory}.
\newblock 2nd edition, Wiley-Interscience, 2006.

\bibitem{tishby1999information}
N.~Tishby, F.~C. Pereira, and W.~Bialek.
\newblock The information bottleneck method.
\newblock In \emph{Proceedings of the 37th Annual Allerton Conference on Communication, Control, and Computing}, pages 368--377, 1999.

\bibitem{hartley2004multiple}
R.~Hartley and A.~Zisserman.
\newblock \emph{Multiple View Geometry in Computer Vision}.
\newblock 2nd edition, Cambridge University Press, 2004.

\end{thebibliography}

\end{document}
